---
title: "Final Hamden Assessed Value Fairness Report"
author: "Lee-Ann Kao, Jake Todd, Ashley Yen"
date: "2025-09-18"
output: pdf_document
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# load libraries
library(dplyr)
library(readxl)
library(rvest)
library(ggplot2)
library(tidyverse)
library(corrplot)
library(MASS)
library(leaflet)
library(RColorBrewer)
library(scales)
library(cv)
```

# Introduction

A Hamden resident who recently purchased his house had his assessed property value significantly increase within a year of the purchase. He began to question whether the assessed value was fair.Our goal is to determine whether the assessed values for residential properties in Hamden, CT are fair and determine if there is any evidence to support contesting the assessed value. 

In order to answer this question, we attempted to recreate the assessor valuation process by using recent 2024 Hamden sales to build a sale pricing model, then applying it to all properties in an attempt to get the would-be 2024 sale price for all homes in Hamden. By comparing our predicted 2024 sale price to the actual assessments in 2023, we determined whether the assessor's values align with our estimate of value and are therefore "fair."

# Reading, Cleaning, and Merging the Data

[______ insert description of how we scraped, cleaned, and merged the data]
```{r, echo=FALSE, warning=FALSE, message=FALSE}
# read in sales data
sale_data <- read_excel("sales_data.xlsx")
sale_data = sale_data %>% 
  filter(Description == 'Single Fam M01') %>%
  rename(
    Sale.Price.2024 = 'Sale Price',
    Sale.Date.2024 = 'Sale Date'
  )

# read in CT property 
hamden_data <- read_csv("Connecticut_CAMA_and_Parcel_Layer_3895368049124948111.csv")
hamden_data = hamden_data %>% 
  filter(Town_Name == "Hamden", Unit_Type == "Single Fam M01")
head(hamden_data)

# read in Hamden html files
for(i in 1:nrow(hamden_data)){
  hamden_data$PID[i] <- str_split(hamden_data$link[i],"-")[[1]][2]
}
hamden_data$PID <- as.numeric(hamden_data$PID)

hamden_data$Sale.Date <- NA
hamden_data$Sale.Price <- NA

hamden_data$PID <- as.numeric(hamden_data$PID)
for(i in 1:nrow(hamden_data)){
  temp_html <- read_html(paste0('Hamden_Sept2025/',hamden_data$PID[i],'.html'))
  tables <- temp_html %>%
    html_elements("table") %>%
    html_table(fill = TRUE)
  if (length(tables) >= 5) {
    sale_table <- tables[[5]]
    sale_table <- sale_table[!sale_table$X1 == "", ]
    sale_table <- sale_table %>%
      pivot_wider(names_from = X1, values_from = X2)
    
    if ("Sale Date" %in% names(sale_table)) {
      hamden_data$Sale.Date[i] <- sale_table$`Sale Date`[1]
    }
    if ("Sale Price" %in% names(sale_table)) {
      hamden_data$Sale.Price[i] <- sale_table$`Sale Price`[1]
    }
  }
}

hamden_data <- hamden_data %>%
  dplyr::select(-'Sale Date', -'Sale Price')

# read in 2009 data
d2009 <- read_csv("Hamden2009.csv")

# left join
sale_data$Location <- paste(sale_data$`Property Number`, sale_data$`Street Name`)
merged <- left_join(hamden_data, sale_data, by = "Location")

colnames(merged) <- gsub(" ", ".", colnames(merged))
head(merged)

# add in centroid/location data
centroids <- readRDS("centroids.no.geometry.rds")
model_data <- merged %>%
  left_join(centroids, by = c("link" = "Link"))

# write csv
write.csv(model_data, "model_data.csv", row.names = FALSE)
```

# Assessed Value/Sale Price Ratio (ASR) Analysis

First, we looked at the Assessed Value/Sale Price ratio (ASR) for homes sold in 2024 to understand if assessed value is considered "fair" in 2024. The histogram and scatterplot below show the distribution of ASR for homes sold in 2024, comparing with 2023 and 2024 assessed values.

The median assessed value over sale price ratio for 2024 is [_____], which is slightly lower than the statutory level of assessed value of 70% for fair market value in Connecticut. Meanwhile, the median ASR for 2023 assessed values is [_____], which is much lower than the statutory level of assessed value of 70% for fair market value in Connecticut. Even accounting for inflation between 2024 and 2023, we see that homes are being severely under-assessed in 2023. The jump in assessed value is most likely because 2023 undervalued homes relative to sale price rather than 2024 overvaluing homes relative to sale price.

From the Leaflet Analysis, we see that almost all ASRs for homes sold in 2024 are below .7. The lowest ASRs are in the Northeast part of Hamden, while the highest ASRs are in the Southeast region. There seems to be more variability in ASR in Southern Hamden, especially compared to the Northwest region, which is predominantly homes that have ASRs of .7 and are most accurately assessed. The second Leaflet plots the 2024 Sales Price of homes

```{r, echo=FALSE}
# read csv
model_data <- read.csv("model_data.csv")

# compute ASR and Equalized Ratio for 2024 sales
model_data <- model_data %>%
  mutate(
    ASR = Current.Year.Assessment / Sale.Price.2024,
    ASR.2023 = Prior.Year.Assessment / Sale.Price.2024,
  )

# Filter to homes sold in 2024
df_2024 <- filter(model_data, !is.na(ASR) & ASR < 2)

# median of data
median(df_2024$ASR, na.rm = TRUE)

# Scatterplot of Assessed Value (2024) vs Actual Sale Price (2024)
ggplot(df_2024, aes(x = Sale.Price.2024, y = Current.Year.Assessment)) +
  geom_point(color = "blue", alpha = 0.5) +
  geom_abline(slope = 0.7, intercept = 0, linetype = "dashed") +
  labs(title = "Assessed Value vs Actual Sale Price (2024)",
       x = "Actual Sale Price",
       y = "Assessed Value")

# Scatterplot of Assessed Value (2023) vs Actual Sale Price (2024)
ggplot(df_2024, aes(x = Sale.Price.2024, y = Prior.Year.Assessment)) +
  geom_point(color = "blue", alpha = 0.5) +
  geom_abline(slope = 0.7, intercept = 0, linetype = "dashed") +
  labs(title = "Assessed Value (2023) vs Actual Sale Price (2024)",
       x = "Actual Sale Price",
       y = "Assessed Value")

# Histogram of ASR 2024
ggplot(df_2024, aes(x = ASR)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white", alpha = 0.8) +
  geom_vline(xintercept = 0.7, color = "red", linetype = "dashed") +
  labs(
    title = "Histogram of ASR for Homes Sold in 2024",
    x = "Assessment-to-Sale Ratio (ASR)",
    y = "Number of Homes"
  )

# Histogram of ASR 2023
ggplot(df_2024, aes(x = ASR.2023)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white", alpha = 0.8) +
  geom_vline(xintercept = 0.7, color = "red", linetype = "dashed") +
  labs(
    title = "Histogram of 2023 ASR for Homes Sold in 2024",
    x = "Assessment-to-Sale Ratio (ASR)",
    y = "Number of Homes"
  )
```

```{r, echo=FALSE}
# Leaflet of ASRs of homes sold in 2024. 
leaflet(df_2024) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~point.x, lat = ~point.y,
    radius = 5,
    color = ~colorNumeric(palette = c("blue","white","red"), domain = df_2024$ASR)(ASR),
    fillOpacity = 0.7,
    popup = ~paste0(Location.y, "<br>ASR = ", round(ASR, 2))
  ) %>%
  addLegend(
    "bottomright", 
    pal = colorNumeric(c("blue","white","red"), df_2024$ASR),
    values = ~ASR,
    title = "ASR",
    opacity = 1
  )
```

```{r, echo=FALSE}
# Leaflet of Sales Prices of homes sold in 2024
# converts sales price 2024 to numeric
sp_raw <- suppressWarnings(as.numeric(df_2024$Sale.Price.2024))
finite <- is.finite(sp_raw)
if (!any(finite)) stop("No finite SalePrice values.")

sp <- sp_raw[finite]

# create quantile bins for sales price, 7 bins
n_bins <- 7
probs  <- seq(0, 1, length.out = n_bins + 1)
bins   <- as.numeric(quantile(sp, probs = probs, na.rm = TRUE))
bins   <- sort(unique(bins)) # drop duplicates
if (length(bins) < 2) bins <- range(sp) # fallback if everything is equal

# uses ColorBrewer palette yellow/orange/red
n_palette <- max(3, min(9, length(bins) - 1))
pal_vec   <- brewer.pal(n_palette, "YlOrRd")

# If bins exceed palette size, interpolate colors
if ((length(bins) - 1) != length(pal_vec)) {
  pal_vec <- colorRampPalette(pal_vec)(length(bins) - 1)
}

# building the bin-based color function
pal_bin <- colorBin(
  palette  = pal_vec,
  bins     = bins,
  domain   = sp_raw,
  right    = TRUE,
  na.color = "transparent"
)

# format legend labels; prettyNum formats numbers with commas
fmt_k <- function(x) paste0("$", prettyNum(round(x, -3), big.mark = ","))
lab_ranges <- paste0(fmt_k(bins[-length(bins)]), " – ", fmt_k(bins[-1]))

# create the leaflet map
leaflet(df_2024) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~point.x, lat = ~point.y,
    radius = 5,
    stroke = TRUE, color = "#8c8c8c", weight = 0.6,
    fill = TRUE, fillColor = ~pal_bin(as.numeric(Sale.Price.2024)), fillOpacity = 0.85,
    popup = ~paste0(
      Location.y, "<br>Sale Price = ",
      ifelse(is.finite(as.numeric(Sale.Price.2024)),
             dollar(round(as.numeric(Sale.Price.2024), -3)),
             "NA")
    )
  ) %>%
  addLegend(
    position = "bottomright",
    colors   = pal_vec,
    labels   = lab_ranges,
    title    = "Sale Price",
    opacity  = 1
  )

```

# Variable Analysis

Next we performed correlation analysis to identify which property features to consider including in our model predicting sale price in 2024. We see that [____ insert] have the strongest correlations with Sale Price. However, there appear to be correlations between these potential predictor variables, and we therefore need to be careful of collinearity when building our model. For example, Living Area is correlated with [_____ insert].

We also looked at scatterplots of all potential variables with Sale Price in order to see if variables meet the linear relationship assumption for linear regression. There appear to be linear relationships between Sale Price and [_____]. However, there may be nonlinear relationships between Sale Price and [_____]. In terms of potential transformations, we see that Sale Price is right-skewed and may benefit from a log transformation.

```{r, echo=FALSE}
# split 
train_data = model_data[!is.na(model_data$`Sale.Price.2024`), ]
test_data = model_data[is.na(model_data$`Sale.Price.2024`), ]

for(i in 1:nrow(train_data)){
  train_data$PID[i] <- str_split(train_data$link[i],"-")[[1]][2]
}
train_data$PID <- as.numeric(train_data$PID)

# limit to potentially relevant variables
train_model_data <- train_data[, c('Land.Acres', 'Living.Area', 'Effective.Area', 'Total.Rooms', 'Number.of.Bedroom', 'Number.of.Baths', 'Sale.Price.2024', 'ayb', 'eyb')]

# make correlation matrix
cor1 <- cor(train_model_data)

corrplot.mixed(cor1, lower.col = "black", upper = "ellipse", tl.col = "black", number.cex = .7, tl.pos = "lt", tl.cex=.7, sig.level = .05)

# make scatterplot matrix
library(psych)
pairs.panels(train_model_data, 
             method = "pearson",
             hist.col = "#00AFBB",
             density = TRUE
             )
```

# Model Building

Next, we built a predictive model using properties that actually sold in 2024 and the variables that we identified above. We performed stepwise regression using AIC and 10-fold cross-validation to select the best model. The final model includes [____ insert] as predictors. Based on cross-validation, there is no sign of significant overfitting, as our full-sample RMSE ([_____]) is similar to the average cross-validated RMSE ([_____]).

We then applied the final model to properties that did not sell in 2024 to predict what those properties would have sold for in 2024. 

Finally, the residual plots below show that the linear model assumptions are reasonably satisfied. The residuals appear to be normally distributed, homoskedastic, and there is no obvious pattern in the residuals vs fitted values plot.

```{r, echo=FALSE}
# stepwise regression using AIC
set.seed(4250)
m.full <- lm(log(Sale.Price.2024) ~ ., data = train_model_data)
m.null <- lm(log(Sale.Price.2024) ~ 1, data = train_model_data)
m.select <- stepAIC(
  m.null,
  direction = "forward",
  trace = FALSE,
  scope = list(lower =  ~ 1, upper = formula(m.full))
)
summary(m.select)

# 10-fold cross-validation for stepwise regression
cv.select <- cv(
  selectStepAIC,
  data = train_model_data,
  seed = 4250,
  working.model = m.null,
  direction = "forward",
  scope = list(lower =  ~ 1, upper = formula(m.full))
)
summary(cv.select)
compareFolds(cv.select)

# final model
fit <- m.select
pred_sale <- predict(fit, newdata = test_data) 
test_data$Sale.Price.2024 <- exp(pred_sale)

# linear model assumptions
plot(fit)
```

# Assessment of Fairness Given the Model
 
The plot of Predicted Price vs Assessed Value of properties shows Assessed Value in red and Predicted Price in blue. The Assessed values are typically lower than both the model-predicted and actual sale prices, suggesting that Hamden homes are being undervalued overall. 

```{r, echo=FALSE}
# values
ggplot(test_data, aes(x = Sale.Price.2024)) +
  geom_point(aes(y = parse_number(as.character(Sale.Price))), color = "blue", alpha = 0.5) +
  geom_point(aes(y = Assessed.Total / 0.70), color = "red", alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(title = "Predicted vs Actual vs Assessed Market Values",
       x = "Predicted Price (Model)",
       y = "Value",
       caption = "Blue = Actual Sale, Red = Assessor (Equalized)")

test_data$Appraised <- test_data$Assessed.Total / 0.70
test_data$residual <- test_data$Sale.Price.2024 - test_data$Appraised
test_data$residual_normalized <- test_data$residual/test_data$Appraised
test_data$residual_normalized

ggplot(data = train_data, aes(x = Assessed.Total, y = Sale.Price.2024)) + geom_point() + geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red")
ggplot(data = test_data, aes(x = Appraised, y = Sale.Price.2024)) + geom_point() + geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red")

```

# Subsets Analysis

The analysis below uses the properties that did not sell in 2024, and therefore we do not have actual sale prices to compare to assessed values. However, we can still analyze the ASR distribution and residuals from our model to assess fairness of assessed values.


From the boxplot, we see that median ASR, using the most recent assessed value for all homes and the 2024 sales price that our model predicted, is less than .5, which is well below the standard of .7. There are several properties with ASRs above 1.0, meaning that either our sales price model is  underestimating the value of some homes, or the assessed value is overestimating. Because homes of higher sales price seem to have higher residuals, we also calculated normalized residual, which is residual divided by the appraised value. 

We plotted the ASR for each property in Hamden on a map, using the property's latitude and longitude to plot. Each property is displayed as a point, with the color indicating the ASR value. Properties with high ASRs (over-assessed) are red, and those with low ASRs (under-assessed) are blue. This helps us determine whether location plays a role in fairness of value. 

The Leaflet plot shows that there is no clear geographic trend for these higher ASRs, although several seem to be in the Northwest region of Hamden. Overall, ASR is well below .7 and fairly consistently blue across the city. 

The following Leaflet plots normalized residuals. There is a clear region of large, positive residuals in the bottom center of Hamden. There also seems to be a cluster of negative residuals in the Northwest region, which is where more expensive homes are. This area also had higher ASRs. We think that our model may be underestimating sales price for more expensive homes. 

```{r}
# boxplot for ASR of Test Data
boxplot(
  test_data$ASR,
  main = "Boxplot of Assessment-to-Sale Ratio (ASR) for Test Data",
  ylab = "ASR (Assessed Value / Sale Price)",
  col = "lightblue",
  border = "darkblue",
  notch = TRUE
)

# Add a reference line at 0.7 (the target)
abline(h = 0.7, col = "red", lwd = 2, lty = 2)

plot(
  test_data$Sale.Price.2024, test_data$residual,
  main = "Residuals vs Sale Price (2024)",
  xlab = "Sale Price (2024, $)",
  ylab = "Residual ($)",
  pch = 19,
  col = rgb(0.2, 0.4, 0.8, 0.5)
)
```

```{r, echo=FALSE}
#Leaflet of ASRs for testing data
test_data <- test_data %>%
  mutate(ASR = Assessed.Total / Sale.Price.2024) %>%
  filter(ASR <= 2, is.finite(ASR))

asr_min <- min(test_data$ASR, na.rm = TRUE)
asr_max <- max(test_data$ASR, na.rm = TRUE)
center  <- 0.7
domain_asr <- c(
  min(asr_min, center - (asr_max - center)), 
  max(asr_max, center + (center - asr_min))
)

pal_asr <- colorNumeric(
  palette = c("blue", "white", "red"),
  domain  = domain_asr,
  na.color = "transparent"
)
leaflet(test_data) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~point.x, lat = ~point.y,
    radius = 5,
    stroke = TRUE, color = "#888888", weight = 0.5,
    fillOpacity = 0.8,
    fillColor = ~pal_asr(ASR),
    popup = ~paste0(Location.y, "<br>ASR = ", round(ASR, 2))
  ) %>%
  addLegend(
    "bottomright",
    pal = pal_asr,
    values = ~ASR,
    title = "Assessment-to-Sale Ratio (ASR)",
    opacity = 1,
    labFormat = labelFormat(digits = 2)
  )
```


```{r, echo=FALSE}
# Leaflet of residuals (test data)
# symmetric trim to remove outliers where sales price is NA/0
res <- test_data$residual_normalized
m   <- quantile(abs(res), 0.95, na.rm = TRUE)
dom <- c(-m, m)

res_trim <- pmin(pmax(res, dom[1]), dom[2])

# build quantile breaks, separately for negative and positive
neg_vals <- res_trim[res_trim < 0]
pos_vals <- res_trim[res_trim > 0]

neg_breaks <- if (length(neg_vals)) quantile(neg_vals, probs = c(0, .25, .5, .75, 1), na.rm = TRUE) else c(-1e-9, 0)
pos_breaks <- if (length(pos_vals)) quantile(pos_vals, probs = c(0, .3, .6, .8, .9, .97, 1), na.rm = TRUE) else c(0, 1e-9)

# Ensure 0 is in the break set exactly once
breaks <- sort(unique(c(neg_breaks, 0, pos_breaks)))

# Palettes per side (multi-hue for positives)
idx0        <- which(breaks == 0)
n_neg_bins  <- idx0 - 1
n_pos_bins  <- (length(breaks) - 1) - n_neg_bins

cols_neg <- if (n_neg_bins > 0) rev(brewer.pal(max(3, n_neg_bins), "Blues"))[seq_len(n_neg_bins)] else character(0)
cols_pos <- if (n_pos_bins > 0)      brewer.pal(max(3, n_pos_bins), "YlOrRd")[seq_len(n_pos_bins)] else character(0)

pal_bin <- colorBin(
  palette  = c(cols_neg, cols_pos),
  bins     = breaks,
  right    = TRUE,
  na.color = "transparent"
)

# Compute colors for each point
cols <- pal_bin(res_trim) # colors for trimmed values
cols[is.na(res)]      <- NA # keep NAs transparent
cols[!is.na(res) & res == 0] <- "#FFFFFF" 

# Leaflet Plot
leaflet(test_data) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~point.x, lat = ~point.y,
    radius = 5,
    stroke = TRUE, color = "#999999", weight = 0.5,
    fill = TRUE, fillColor = cols,  # <-- no tilde
    fillOpacity = 0.8,
    popup = ~paste0(
      Location.y, "<br>Normalized Residual = ",
      ifelse(is.finite(residual_normalized),
             round(residual_normalized, 2),
             "NA")
    )
  ) %>%
  addLegend(
    position = "bottomright",
    pal      = pal_bin,
    values   = res_trim,
    title    = "Normalized Residual",
    opacity  = 1,
    labFormat = labelFormat(
      transform = identity,
      digits    = 2
    )
  ) %>%
  addLegend(
    "bottomright",
    colors  = "#FFFFFF",
    labels  = "0",
    opacity = 1,
    title   = NULL
  )
```


```{r, echo=FALSE}
# Data prep: coerce to numeric & drop NAs
df <- test_data
df$SalePrice_num <- suppressWarnings(as.numeric(df$Sale.Price.2024))
df$residual_num  <- suppressWarnings(as.numeric(df$residual))
df2 <- subset(df, is.finite(SalePrice_num) & is.finite(residual_num))

# Correlation stats
ct <- cor.test(df2$SalePrice_num, df2$residual_num)
ann_txt <- sprintf("Pearson r = %.3f\np = %.3g\nn = %d", ct$estimate, ct$p.value, nrow(df2))

# Scatter with linear fit
p <- ggplot(df2, aes(x = SalePrice_num, y = residual_num)) +
  geom_point(alpha = 0.35, size = 1.6) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.9, color = "firebrick") +
  annotate("text", x = Inf, y = Inf, label = ann_txt, hjust = 1.02, vjust = 1.1, size = 3.6) +
  scale_x_continuous(labels = label_dollar(accuracy = 1)) +
  labs(
    x = "Sale Price",
    y = "Residual",
    title = "Residuals vs. Sale Price",
  ) +
  theme_minimal(base_size = 12)

print(p)
```

# Conclusion

Our analysis shows that assessed value in 2023 is significantly lower than sale price, indicating that homes are being undervalued overall in 2023. This could account for the apparent jump in assessed value that our homeowner notcied. The model we built using 2024 sales data suggests that the assessor's 2023 values are much lower than our estimates of 2024 market value, further supporting the idea that assessed values may not be fair (although opposite what the homeowner suspected).

Through our model, we found that certain property features, such as [____ insert], are strong predictors of sale price and are likely included in the assessor's model. The residual analysis indicates that the model assumptions are reasonably satisfied, and there is no significant overfitting based on cross-validation results. However, there are limitations to our analysis. Our model is based on the assumption that the assessor uses a similar set of variables and a linear regression approach, which may not be the case. 

Additionally, this model does not account for all possible factors that could influence property values, such as neighborhood effects or recent renovations. This is apparent in the map analysis where we see clusters of over-assessed and under-assessed homes, suggesting that location may play a role in fairness of value.
