---
title: "Final-Hamden-assessed-value-fairness"
author: "Lee-Ann Kao, Jake Todd, Ashley Yen"
date: "2025-09-18"
output: pdf_document
---

```{r}
# load libraries
library(dplyr)
library(readxl)
library(ggplot2)
library(tidyverse)
library(corrplot)
library(MASS)
library(leaflet)
library(RColorBrewer)
library(scales)
```

# Introduction

A Hamden homeowner saw a large assessment increase shortly after purchasing and wanted to know whether the new assessed value is fair. In order to answer this question, we attempted to recreate the assessor valuation process by using recent 2024 sales to build a sale pricing model, then applying it to all properties in an attempt to get the would-be 2024 sale price for all homes. By comparing our predicted 2024 sale price to the actual assessments, we determined whether the assessor's values align with our estimate of value and are therefore "fair." We define fair as [____ insert definition of fair].

# Reading, Cleaning, and Merging the Data

[______ insert description of how we scraped, cleaned, and merged the data]

```{r}
model_data = read_csv("data.csv")
```

# Assessed Value/Sale Price Ratio (ASR) Analysis

First, we looked at the Assessed Value/Sale Price ratio (ASR) for homes sold in 2024 to understand if assessed value is considered "fair" in 2024. The histogram and scatterplot below show the distribution of ASR for homes sold in 2024. 

The median assessed value over sale price ratio is [_____], which is much lower than the statutory level of assessed value of 70% for fair market value in Connecticut. Even accounting for inflation between 2024 and 2023, we see that homes are being severely under-assessed in 2023.

```{r}
# Filter to homes sold in 2024
df_2024 <- filter(model_data, !is.na(Sale_2024) & Sale_2024 > 0 & ASR < 2)

# median of data
median(df_2024$ASR, na.rm = TRUE)

# Scatterplot of Assessed Value (2023) vs Actual Sale Price (2024)
ggplot(df_2024, aes(x = Sale_2024, y = Assessed.Total)) +
  geom_point(color = "blue", alpha = 0.5) +
  geom_abline(slope = 0.7, intercept = 0, linetype = "dashed") +
  labs(title = "Assessed Value (2023) vs Actual Sale Price (2024)",
       x = "Actual Sale Price (2024)",
       y = "Assessed Value (2023)")

# Histogram of ASR
ggplot(df_2024, aes(x = ASR)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white", alpha = 0.8) +
  geom_vline(xintercept = 0.7, color = "red", linetype = "dashed") +
  labs(
    title = "Histogram of ASR for Homes Sold in 2024",
    x = "Assessment-to-Sale Ratio (ASR)",
    y = "Number of Homes"
  )
```
Leaflet of ASRs of homes sold in 2024. 
```{r}
leaflet(df_2024) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~point.x, lat = ~point.y,
    radius = 5,
    color = ~colorNumeric(palette = c("blue","white","red"), domain = df_2024$ASR)(ASR),
    fillOpacity = 0.7,
    popup = ~paste0(Location.y, "<br>ASR = ", round(ASR, 2))
  ) %>%
  addLegend(
    "bottomright", 
    pal = colorNumeric(c("blue","white","red"), df_2024$ASR),
    values = ~ASR,
    title = "ASR",
    opacity = 1
  )
```

Leaflet of Sales Prices of homes sold in 2024
```{r}
# --- 1) Numeric & finite values ---
sp_raw <- suppressWarnings(as.numeric(df_2024$Sale_2024))
finite <- is.finite(sp_raw)
if (!any(finite)) stop("No finite SalePrice values.")

sp <- sp_raw[finite]

# --- 2) Quantile bins
n_bins <- 7
probs  <- seq(0, 1, length.out = n_bins + 1)
bins   <- as.numeric(quantile(sp, probs = probs, na.rm = TRUE))
bins   <- sort(unique(bins))                  # drop duplicates from ties
if (length(bins) < 2) bins <- range(sp)       # fallback if everything is equal

n_palette <- max(3, min(9, length(bins) - 1))
pal_vec   <- brewer.pal(n_palette, "YlOrRd")

# If our bins exceed palette size (rare after unique), interpolate colors
if ((length(bins) - 1) != length(pal_vec)) {
  pal_vec <- colorRampPalette(pal_vec)(length(bins) - 1)
}

# --- 3) Color function over bins ---
pal_bin <- colorBin(
  palette  = pal_vec,
  bins     = bins,
  domain   = sp_raw,          # domain can include NAs; colorBin only uses bins
  right    = TRUE,
  na.color = "transparent"
)

# --- 4) Legend labels as ranges (rounded to nearest thousand, with commas) ---
fmt_k <- function(x) paste0("$", prettyNum(round(x, -3), big.mark = ","))
lab_ranges <- paste0(fmt_k(bins[-length(bins)]), " – ", fmt_k(bins[-1]))

leaflet(df_2024) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~point.x, lat = ~point.y,
    radius = 5,
    stroke = TRUE, color = "#8c8c8c", weight = 0.6,
    fill = TRUE, fillColor = ~pal_bin(as.numeric(Sale_2024)), fillOpacity = 0.85,
    popup = ~paste0(
      Location.y, "<br>Sale Price = ",
      ifelse(is.finite(as.numeric(Sale_2024)),
             dollar(round(as.numeric(Sale_2024), -3)),
             "NA")
    )
  ) %>%
  # Use manual legend so we can show explicit ranges per bin
  addLegend(
    position = "bottomright",
    colors   = pal_vec,
    labels   = lab_ranges,
    title    = "Sale Price (quantile bins)",
    opacity  = 1
  )

```

# Variable Analysis

Next we performed correlation analysis to identify which property features to consider including in our model predicting sale price in 2024.

We see that [____ insert] have the strongest correlations with Sale Price. However, there appear to be correlations between these potential predictor variables, and we therefore need to be careful of collinearity when building our model. For example, Living Area is correlated with [_____ insert].

In terms of potential transformations, we see that Sale Price is right-skewed and may benefit from a log transformation.

There appear to be linear relationships between Sale Price and [_____]. However, there may be nonlinear relationships between Sale Price and [_____].

```{r}
# split 
for(i in 1:nrow(model_data)){
  model_data$Sale.Year[i] <- substr(model_data$Sale.Date[i], nchar(model_data$Sale.Date[i]) - 3, nchar(model_data$Sale.Date[i]))
}
model_data$Sale.Year <- as.numeric(model_data$Sale.Year)

train_data = model_data[!is.na(model_data$`Sale_2024`), ]
test_data = model_data[is.na(model_data$`Sale_2024`), ]

for(i in 1:nrow(train_data)){
  train_data$PID[i] <- str_split(train_data$link[i],"-")[[1]][2]
}
train_data$PID <- as.numeric(train_data$PID)

# limit to potentially relevant variables
train_model_data <- train_data[, c('Land.Acres', 'Living.Area', 'Effective.Area', 'Total.Rooms', 'Number.of.Bedroom', 'Number.of.Baths', 'Sale_2024', 'ayb', 'eyb')]

# make correlation matrix
cor1 <- cor(train_model_data)

corrplot.mixed(cor1, lower.col = "black", upper = "ellipse", tl.col = "black", number.cex = .7, tl.pos = "lt", tl.cex=.7, sig.level = .05)

# make scatterplot matrix
library(psych)
pairs.panels(train_model_data, 
             method = "pearson",
             hist.col = "#00AFBB",
             density = TRUE
             )
```

# Model Building

Next, we built a predictive model using properties that actually sold in 2024. We then applied the model to properties that did not sell in 2024 in order to obtain an estimate of what those properties would have sold for in 2024. 

We performed stepwise regression using AIC and 10-fold cross-validation to select the best model. The final model includes [____ insert] as predictors. Based on cross-validation, there is no sign of significant overfitting, as our full-sample RMSE ([_____]) is similar to the average cross-validated RMSE ([_____]).

Finally, the residual plots below show that the linear model assumptions are reasonably satisfied. The residuals appear to be normally distributed, homoskedastic, and there is no obvious pattern in the residuals vs fitted values plot.

```{r}
# stepwise regression using AIC
set.seed(4250)
m.full <- lm(log(Sale_2024) ~ ., data = train_model_data)
m.null <- lm(log(Sale_2024) ~ 1, data = train_model_data)
m.select <- stepAIC(
  m.null,
  direction = "forward",
  trace = FALSE,
  scope = list(lower =  ~ 1, upper = formula(m.full))
)
summary(m.select)

# 10-fold cross-validation for stepwise regression
cv.select <- cv(
  selectStepAIC,
  data = train_model_data,
  seed = 4250,
  working.model = m.null,
  direction = "forward",
  scope = list(lower =  ~ 1, upper = formula(m.full))
)
summary(cv.select)
compareFolds(cv.select)

# final model
fit <- m.select
pred_sale <- predict(fit, newdata = test_data) 
test_data$Sale_2024 <- exp(pred_sale)

# linear model assumptions
plot(fit)
```

```{r}
# make correlation matrix
cor1 <- cor(train_data[, c('Sale_2024', 'Pre.Yr.Assessed.Total', 'Land.Acres', 'Living.Area', 'Effective.Area', 'Total.Rooms', 'Number.of.Bedroom', 'Number.of.Baths', 'Number.of.Half.Baths', 'ayb', 'eyb', 'appraised_2009')], use = "pairwise.complete.obs")

corrplot.mixed(cor1, lower.col = "black", upper = "ellipse", tl.col = "black", number.cex = .7, tl.pos = "lt", tl.cex=.7, sig.level = .05)

# limit to potentially relevant variables
train_model_data <- train_data[, c('Land.Acres', 'Living.Area', 'Effective.Area', 'Total.Rooms', 'Number.of.Bedroom', 'Number.of.Baths', 'Sale_2024', 'ayb', 'eyb', 'appraised_2009')]
train_model_data = train_model_data[!is.na(train_model_data$appraised_2009),]

# perform best subsets regression
full_model <- lm(log(Sale_2024) ~ ., data = train_model_data)
null_model <- lm(log(Sale_2024) ~ 1, data = train_model_data)
n <- nrow(train_model_data)

# stepwise regression using BIC
stepwise_bic_model <- stepAIC(null_model,
                              direction = "both",
                              scope = list(upper = full_model, lower = null_model),
                              k = log(n))
summary(stepwise_bic_model)

# stepwise regression AIC
stepwise_aic_model <- stepAIC(full_model, direction = "both", trace = FALSE)
summary(stepwise_aic_model)

# final model
fit <- stepwise_aic_model
pred_sale <- predict(fit, newdata = test_data) 
test_data$Sale_2024 <- exp(pred_sale)
```

# Assessment of Fairness Given the Model

The plot of Predicted Price vs Assessed Value of properties shows Assessed Value in red and Predicted Price in blue. The Assessed values are typically lower than both the model-predicted and actual sale prices, suggesting that Hamden homes are being undervalued overall. 

In addition, we plotted the ASR for each property in Hamden on a map, using the property's latitude and longitude to plot. Each property is displayed as a point, with the color indicating the ASR value. Properties with high ASRs (over-assessed) are red, and those with low ASRs (under-assessed) are blue. This helps us determine whether location plays a role in fairness of value. 

```{r}
# values
ggplot(test_data, aes(x = Sale_2024)) +
  geom_point(aes(y = Sale.Price), color = "blue", alpha = 0.5) +
  geom_point(aes(y = Assessed.Total / 0.70), color = "red", alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(title = "Predicted vs Actual vs Assessed Market Values",
       x = "Predicted Price (Model)",
       y = "Value",
       caption = "Blue = Actual Sale, Red = Assessor (Equalized)")

test_data$Appraised <- test_data$Assessed.Total / 0.70
test_data$residual <- test_data$Sale_2024 - test_data$Appraised
test_data$residual_normalized <- test_data$residual/test_data$Appraised
test_data$residual_normalized

ggplot(data = train_data, aes(x = `Current Year Appraisal`, y = Sale_2024)) + geom_point() + geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red")
ggplot(data = test_data, aes(x = Appraised, y = Sale_2024)) + geom_point() + geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red")
```

# Subsets Analysis

```{r}
leaflet(model_data) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~point.x, lat = ~point.y,
    radius = 5,
    color = ~colorNumeric(palette = c("blue","white","red"), domain = model_data$ASR)(ASR),
    fillOpacity = 0.7,
    popup = ~paste0(Location.y, "<br>ASR = ", round(ASR, 2))
  ) %>%
  addLegend(
    "bottomright", 
    pal = colorNumeric(c("blue","white","red"), model_data$ASR),
    values = ~ASR,
    title = "ASR",
    opacity = 1
  )
```


```{r}
library(leaflet)
library(RColorBrewer)

# --- 1) symmetric trim to remove outliers where sales price is NA/0---
res <- test_data$residual
m   <- quantile(abs(res), 0.95, na.rm = TRUE)
dom <- c(-m, m)

res_trim <- pmin(pmax(res, dom[1]), dom[2])

# --- 2) Build quantile breaks to help with color scale ---
neg_vals <- res_trim[res_trim < 0]
pos_vals <- res_trim[res_trim > 0]

neg_breaks <- if (length(neg_vals)) quantile(neg_vals, probs = c(0, .25, .5, .75, 1), na.rm = TRUE) else c(-1e-9, 0)
pos_breaks <- if (length(pos_vals)) quantile(pos_vals, probs = c(0, .3, .6, .8, .9, .97, 1), na.rm = TRUE) else c(0, 1e-9)

# Ensure 0 is in the break set exactly once
breaks <- sort(unique(c(neg_breaks, 0, pos_breaks)))

# --- 3) Palettes per side (multi-hue for positives) ---
idx0        <- which(breaks == 0)
n_neg_bins  <- idx0 - 1
n_pos_bins  <- (length(breaks) - 1) - n_neg_bins

cols_neg <- if (n_neg_bins > 0) rev(brewer.pal(max(3, n_neg_bins), "Blues"))[seq_len(n_neg_bins)] else character(0)
cols_pos <- if (n_pos_bins > 0)      brewer.pal(max(3, n_pos_bins), "YlOrRd")[seq_len(n_pos_bins)] else character(0)

pal_bin <- colorBin(
  palette  = c(cols_neg, cols_pos),
  bins     = breaks,
  right    = TRUE,
  na.color = "transparent"
)

# --- 4) Compute colors ---
cols <- pal_bin(res_trim)             # colors for trimmed values
cols[is.na(res)]      <- NA           # keep NAs transparent
cols[!is.na(res) & res == 0] <- "#FFFFFF" 

# --- 5) Plot ---
leaflet(test_data) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~point.x, lat = ~point.y,
    radius = 5,
    stroke = TRUE, color = "#999999", weight = 0.5,
    fill = TRUE, fillColor = ~cols, fillOpacity = 0.8,
    popup = ~paste0(Location.y, "<br>residual = ", round(residual, 2))
  ) %>%
  addLegend(
  position = "bottomright",
  pal      = pal_bin,
  values   = res_trim,
  title    = "Residual",
  opacity  = 1,
  labFormat = labelFormat(
    transform = function(x) round(x, -3),          # round to nearest 1000
    big.mark  = ",",                               # add commas
    digits    = 0                                  # no decimals
  )
  ) %>%
  addLegend(
    "bottomright",
    colors  = "#FFFFFF", labels = "0", opacity = 1, title = NULL
  )



```


```{r}
# Data prep: coerce to numeric & drop NAs
df <- test_data
df$SalePrice_num <- suppressWarnings(as.numeric(df$Sale_2024))
df$residual_num  <- suppressWarnings(as.numeric(df$residual))
df2 <- subset(df, is.finite(SalePrice_num) & is.finite(residual_num))

# Correlation stats
ct <- cor.test(df2$SalePrice_num, df2$residual_num)
ann_txt <- sprintf("Pearson r = %.3f\np = %.3g\nn = %d", ct$estimate, ct$p.value, nrow(df2))

# Scatter with linear fit
p <- ggplot(df2, aes(x = SalePrice_num, y = residual_num)) +
  geom_point(alpha = 0.35, size = 1.6) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.9, color = "firebrick") +
  annotate("text", x = Inf, y = Inf, label = ann_txt, hjust = 1.02, vjust = 1.1, size = 3.6) +
  scale_x_continuous(labels = label_dollar(accuracy = 1)) +
  labs(
    x = "Sale Price",
    y = "Residual",
    title = "Residuals vs. Sale Price",
  ) +
  theme_minimal(base_size = 12)

print(p)

```

# Conclusion
Our analysis shows that homes are generally being undervalued relative to their sales prices. From a homeowner's perspective, this is variable as it yields them a lower property tax bill. 

1. just look at properties sold in 2024 - create leaflet showing the ASRs and which are over/under valued
2. look at sales prices and see if there is a trend in the subsets

- percent difference

- plot scatterplot of residuals vs something else

